---
title: "Credit Risk Modeling & Gender Fairness in Loan Approvals"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5)
```

## Overview

This project builds a logistic regression model to predict credit risk using the German Credit dataset, then examines whether a single decision threshold treats male and female applicants fairly and what happens when you adjust for that.

The dataset has 1,000 loan applicants with features like loan duration, credit history, savings, employment status, and age. The target variable is `credit_risk` (1 = good, 0 = bad).

---

## Data

```{r load-data}
data <- read.csv("German Credit Data.csv")
dim(data)
str(data)
```

```{r gender-summary}
gender_summary <- data.frame(
  Gender = c("Male", "Female"),
  Count = c(sum(data$gender == "Male"), sum(data$gender == "Female")),
  Avg_Credit_Risk = c(
    round(mean(data$credit_risk[data$gender == "Male"]), 3),
    round(mean(data$credit_risk[data$gender == "Female"]), 3)
  )
)
knitr::kable(gender_summary, caption = "Gender breakdown and average credit risk")
```

Males make up 69% of the dataset and have a slightly higher average credit risk score (0.723 vs 0.648). This imbalance matters because it will show up in how the model behaves across groups.

---

## Model

60/40 train/test split, then logistic regression using all features.

```{r model}
set.seed(1)
trainindex <- sample(1000, 600)
train <- data[trainindex, ]
test  <- data[-trainindex, ]

model <- glm(credit_risk ~ ., data = train, family = binomial)
summary(model)
```

### Significant predictors

```{r coef-plot}
library(ggplot2)

coef_df <- as.data.frame(summary(model)$coefficients)
coef_df$Variable <- rownames(coef_df)
coef_df <- coef_df[coef_df$Variable != "(Intercept)", ]
coef_df$Significant <- coef_df$`Pr(>|z|)` < 0.05
coef_df <- coef_df[order(abs(coef_df$Estimate), decreasing = TRUE), ]
coef_df <- head(coef_df, 20)

ggplot(coef_df, aes(x = reorder(Variable, Estimate), y = Estimate, fill = Significant)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = "grey70", "TRUE" = "#2c7bb6")) +
  labs(title = "Top 20 Logistic Regression Coefficients",
       x = NULL, y = "Coefficient Estimate",
       fill = "Significant (p < 0.05)") +
  theme_minimal()
```

---

## Baseline performance (single threshold = 0.5)

```{r baseline}
train$pred_prob   <- predict(model, newdata = train, type = "response")
predicted.class   <- (train$pred_prob >= 0.5) * 1
actual.class      <- train$credit_risk

cm <- table(Actual = actual.class, Predicted = predicted.class)
knitr::kable(cm, caption = "Confusion matrix, training set, threshold = 0.5")

overall.error <- sum(actual.class != predicted.class) / nrow(train)
cat("Overall error rate:", round(overall.error, 4))
```

---

## Fairness analysis: does the model treat genders equally?

At the same 0.5 threshold, we can check whether males and females are approved at similar rates and whether the model is equally accurate for both groups.

```{r fairness-baseline}
PR       <- sum(predicted.class == 1) / nrow(train)
PR_Male  <- sum(predicted.class[train$gender == "Male"] == 1)  / sum(train$gender == "Male")
PR_Female<- sum(predicted.class[train$gender == "Female"] == 1)/ sum(train$gender == "Female")

TPR      <- sum(predicted.class == 1 & actual.class == 1) / sum(actual.class == 1)
TPR_Male <- sum(predicted.class[train$gender == "Male"] == 1 & actual.class[train$gender == "Male"] == 1) /
            sum(actual.class[train$gender == "Male"] == 1)
TPR_Female <- sum(predicted.class[train$gender == "Female"] == 1 & actual.class[train$gender == "Female"] == 1) /
              sum(actual.class[train$gender == "Female"] == 1)

fairness_table <- data.frame(
  Group    = c("Overall", "Male", "Female"),
  Approval_Rate_PR  = round(c(PR, PR_Male, PR_Female), 3),
  True_Positive_Rate = round(c(TPR, TPR_Male, TPR_Female), 3)
)
knitr::kable(fairness_table, caption = "Approval rate and TPR by gender at threshold = 0.5")
```

```{r fairness-bar}
library(tidyr)

fairness_long <- pivot_longer(fairness_table, cols = c(Approval_Rate_PR, True_Positive_Rate),
                              names_to = "Metric", values_to = "Value")

ggplot(fairness_long, aes(x = Group, y = Value, fill = Group)) +
  geom_bar(stat = "identity", width = 0.6) +
  facet_wrap(~Metric, labeller = as_labeller(c(
    Approval_Rate_PR   = "Approval Rate (PR)",
    True_Positive_Rate = "True Positive Rate (TPR)"
  ))) +
  scale_fill_manual(values = c("Overall" = "grey60", "Male" = "#2c7bb6", "Female" = "#d7191c")) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(title = "Gender disparity at a single 0.5 threshold",
       x = NULL, y = NULL, fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The gap is visible. Males get approved at 80% vs females at 68%, and the TPR gap is about 5 percentage points.

---

## Predicted probability distributions by gender

Before optimizing thresholds, it helps to see why the disparity exists. The model produces different probability distributions for males and females.

```{r density-plot}
ggplot(train, aes(x = pred_prob, fill = gender, color = gender)) +
  geom_density(alpha = 0.3, linewidth = 1) +
  scale_fill_manual(values  = c("Male" = "#2c7bb6", "Female" = "#d7191c")) +
  scale_color_manual(values = c("Male" = "#2c7bb6", "Female" = "#d7191c")) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "black") +
  labs(title = "Predicted probability distributions by gender",
       subtitle = "Dashed line = 0.5 threshold",
       x = "Predicted Probability of Good Credit", y = "Density",
       fill = NULL, color = NULL) +
  theme_minimal()
```

The female distribution is shifted left, meaning the model assigns lower probabilities to females on average. A single cutoff at 0.5 therefore catches fewer females in the approved bucket, even among genuinely creditworthy ones.

---

## Optimizing thresholds for fairness

The goal is to find a pair of thresholds (one for males, one for females) where TPR_Male and TPR_Female are within 0.1% of each other while keeping overall error as low as possible.

```{r threshold-optimization}
TPR_Male_vec   <- matrix(0, 1, 100)
TPR_Female_vec <- matrix(0, 1, 100)
error_Male_vec <- matrix(0, 1, 100)
error_Female_vec <- matrix(0, 1, 100)

for (c in 1:100) {
  cut_off <- c / 100
  pred_m  <- (train$pred_prob[train$gender == "Male"]   >= cut_off) * 1
  pred_f  <- (train$pred_prob[train$gender == "Female"] >= cut_off) * 1
  TPR_Male_vec[c]    <- sum(pred_m == 1 & actual.class[train$gender == "Male"]   == 1) / sum(actual.class[train$gender == "Male"]   == 1)
  TPR_Female_vec[c]  <- sum(pred_f == 1 & actual.class[train$gender == "Female"] == 1) / sum(actual.class[train$gender == "Female"] == 1)
  error_Male_vec[c]  <- sum(pred_m != actual.class[train$gender == "Male"])
  error_Female_vec[c]<- sum(pred_f != actual.class[train$gender == "Female"])
}

min_overall.error <- 1
thresholds_pair   <- c(0, 0)

for (c_m in 1:100) {
  for (c_f in 1:100) {
    if (abs(TPR_Male_vec[c_m] - TPR_Female_vec[c_f]) < 0.001) {
      overall.error <- (error_Male_vec[c_m] + error_Female_vec[c_f]) / nrow(train)
      if (overall.error < min_overall.error) {
        min_overall.error <- overall.error
        thresholds_pair   <- c(c_m / 100, c_f / 100)
      }
    }
  }
}

cat("Optimal thresholds - Male:", thresholds_pair[1], "| Female:", thresholds_pair[2])
cat("\nMinimum overall error:", round(min_overall.error, 4))
```

```{r tpr-curve}
tpr_df <- data.frame(
  Threshold = rep((1:100)/100, 2),
  TPR       = c(as.vector(TPR_Male_vec), as.vector(TPR_Female_vec)),
  Gender    = rep(c("Male", "Female"), each = 100)
)

ggplot(tpr_df, aes(x = Threshold, y = TPR, color = Gender)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = thresholds_pair[1], linetype = "dashed", color = "#2c7bb6") +
  geom_vline(xintercept = thresholds_pair[2], linetype = "dashed", color = "#d7191c") +
  scale_color_manual(values = c("Male" = "#2c7bb6", "Female" = "#d7191c")) +
  scale_x_continuous(breaks = seq(0, 1, 0.1)) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "TPR vs threshold for each gender",
       subtitle = "Dashed lines mark the optimal thresholds where TPR equalizes",
       x = "Threshold", y = "True Positive Rate", color = NULL) +
  theme_minimal()
```

---

## Test set evaluation with optimized thresholds

```{r test-eval}
test$pred_prob <- predict(model, newdata = test, type = "response")
test.actual.class <- test$credit_risk

pred_male   <- (test$pred_prob[test$gender == "Male"]   >= thresholds_pair[1]) * 1
pred_female <- (test$pred_prob[test$gender == "Female"] >= thresholds_pair[2]) * 1

err_m <- sum(pred_male   != test.actual.class[test$gender == "Male"])
err_f <- sum(pred_female != test.actual.class[test$gender == "Female"])
test_error <- (err_m + err_f) / nrow(test)

# Fairness metrics on test set
test_PR_Male   <- sum(pred_male   == 1) / sum(test$gender == "Male")
test_PR_Female <- sum(pred_female == 1) / sum(test$gender == "Female")
test_TPR_Male  <- sum(pred_male   == 1 & test.actual.class[test$gender == "Male"]   == 1) / sum(test.actual.class[test$gender == "Male"]   == 1)
test_TPR_Female<- sum(pred_female == 1 & test.actual.class[test$gender == "Female"] == 1) / sum(test.actual.class[test$gender == "Female"] == 1)

results_table <- data.frame(
  Setup = c("Single threshold (0.5)", "Optimized thresholds (0.42 / 0.34)"),
  Male_Threshold   = c(0.50, thresholds_pair[1]),
  Female_Threshold = c(0.50, thresholds_pair[2]),
  Test_Error       = c(NA,   round(test_error, 4))
)
knitr::kable(results_table, caption = "Threshold comparison")

cat("Test set overall error with optimized thresholds:", round(test_error, 4))
```

```{r final-fairness}
final_fairness <- data.frame(
  Group    = c("Male", "Female"),
  Approval_Rate = round(c(test_PR_Male, test_PR_Female), 3),
  TPR           = round(c(test_TPR_Male, test_TPR_Female), 3)
)
knitr::kable(final_fairness, caption = "Fairness metrics on test set with optimized thresholds")
```

---

## Summary

| | Single threshold (0.5) | Optimized thresholds |
|---|---|---|
| Male threshold | 0.50 | 0.42 |
| Female threshold | 0.50 | 0.34 |
| TPR gap (Male - Female) | ~5 pp | < 0.1 pp |
| Overall error (train) | 20.7% | 20.8% |
| Overall error (test) | not evaluated | 24.8% |

Using group-specific thresholds closes the TPR gap almost entirely at a cost of less than 0.2 percentage points in overall accuracy. The tradeoff is small and the fairness gain is significant.

This approach is sometimes called **equalized opportunity**: equalizing TPR across groups rather than using a single threshold that inadvertently advantages one group due to distributional differences in the model's outputs.
